---
title: "Parkinson data - I"
author: "Giovanni Saraceno"
output:
  pdf_document:
    keep_tex: true
    toc: true
  html_document:
    df_print: paged
    toc: true
date: ''
bibliography: ../notes.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

We consider the data set related to the study in @tsanas2009accurate. 
Data can be found at [https:// archive.ics.uci.edu/ ml/ datasets/ Parkinsons+Telemonitoring](https:// archive.ics.uci.edu/ ml/ datasets/ Parkinsons+Telemonitoring). 

We start loading the data.
```{r}
dat <- read.csv('parkinsons_updrs.data', he = TRUE, sep = ',')
str(dat)
```

#### Description of the variables

- subject# - Integer that uniquely identifies each subject
- `age` - Subject age
- `sex` - Subject gender ‘0’ - male, ‘1’ - female
- `test_time` - Time since recruitment into the trial. The integer part is the number of days since
recruitment.
- `motor_UPDRS` - Clinician’s motor UPDRS score, linearly interpolated
- `total_UPDRS` - Clinician’s total UPDRS score, linearly interpolated
- `Jitter`(%), `Jitter(Abs)`, `Jitter:RAP`, `Jitter:PPQ5`, `Jitter:DDP` - Several measures of variation in fundamental frequency
- `Shimmer`, `Shimmer(dB)`, `Shimmer:APQ3`, `Shimmer:APQ5`, `Shimmer:APQ11`, `Shimmer:DDA` - Several
measures of variation in amplitude 
- `NHR`, `HNR` - Two measures of ratio of noise to tonal components in the voice
- `RPDE` - A nonlinear dynamical complexity measure
- `DFA` - Signal fractal scaling exponent
- `PPE` - A nonlinear measure of fundamental frequency variation

```{r}
str(dat)
```

#### Data cleaning

```{r}
dat$subject. <- as.factor(dat$subject.)
dat$sex <- ifelse(dat$sex == 1, 'F', 'M')
dat$sex <- as.factor(dat$sex)
```

#### Descriptive analysis

Let’s have a look at a brief descriptive summary of all the variables
```{r}
summary(dat)
```

We want to check whether NAs are present.
```{r}
colSums(is.na(dat))
```

We can also check the variance of each variable, in order to look for ‘almost’ constant variable (variance very
close to zero).
```{r}
apply(dat, 2, function(x) var(x))
```

We can now investigates the variables in details. We can start with the categorical variables and in relation to `total_UPDRS` since it is our variable of interest. \
We start with `subject`
```{r}
ggplot(dat, aes(x=subject.)) +
  geom_bar() + 
  theme_minimal()
```

We have different number of measurements fo the subjects. 
```{r}
ggplot(dat, aes(x = subject., y = total_UPDRS)) +
  geom_point() + 
  theme_minimal()
```

It is useful also to explore the sex variable
```{r}
dat %>% group_by(sex) %>% summarise(n = n())
ggplot(dat, aes(x = sex, y = total_UPDRS)) +
  geom_boxplot() + 
  theme_minimal()
```

Now let's look at *age*
```{r}
dat %>% group_by(age) %>% summarise(n = n())
ggplot(dat, aes(x = age)) +
  geom_bar() + 
  theme_minimal()
ggplot(dat, aes(x = age, y = total_UPDRS)) +
  geom_point() + 
  theme_minimal()
```


There is also the variable that indicates the time at which the data are collected for each subject. Let us look at the values of `total_UPDRS` in time, grouped by subject and colored by sex
```{r}
ggplot(dat, aes(x = test_time, y = total_UPDRS, group = subject., col=sex)) + 
  geom_line() +
  geom_point() +
  theme_minimal()
```
or by age
```{r}
ggplot(dat, aes(x = test_time, y = total_UPDRS, group = subject., col=age)) + 
  geom_line() +
  geom_point() +
  theme_minimal()
```

Consider now the covariates
```{r, message=FALSE}
library(reshape2)
dat_dens <- melt(dat[, 7:22])
ggplot(dat_dens, aes(x = value)) + 
  geom_histogram(aes(y=..density..), bins=40) +
  geom_density(color="red", linetype="dashed") +
  facet_wrap(~ variable, ncol = 4, scales = 'free') +
  theme_bw() +
  xlab("")
```

We can see that `HNR` and `RPDE` are bell-shaped, while `DFA` seems to be bi-modal. \
We can continue computing the correlation of these variables with the `total_UPDRS`
```{r}
apply(dat[, 7:22], 2, function(x) cor(dat$total_UPDRS, x, method = 'spearman'))
```
and the corresponding scatter plots
```{r}
dat_dens$y <- rep(dat$total_UPDRS, length(7:22))
ggplot(dat_dens, aes(x = value, y = y)) + 
  geom_point(size = 1) +
  facet_wrap(~ variable, ncol = 4, scales = 'free') +
  theme_bw()
```

Finally we compute the correlations among the covariates
```{r, warning=FALSE}
cor_cov <- cor(dat[, 7:22], method = 'spearman')
library(corrplot)
corrplot(cor_cov)
corrplot.mixed(cor_cov, tl.col = 'black', insig = 'blank', tl.pos = 'lt', number.cex = 0.5)
```


#### References
